sdmTMB: Fast, flexible, and user-friendly spatial and spatiotemporal generalized linear mixed-effects models

### 2021-11-18 doc/paper.Rmd
Develop paper

sdmTMB fits spatially explicit geostatistical GLMMs to point data.
Point data simply means the collected data have explicit spatial coordinates (vs. areal data that might represent polygons or grids).
The process component of an sdmTMB model can be formed by any combination of main ('fixed') effects, spatial random fields, spatiotemporal random fields, IID random intercepts, time-varying effects, and spatially varying effects (Fig. \@ref(fig:diagram)).
This process component is combined with an observation error family (e.g., Gaussian, Gamma, binomial, Tweedie) and link (e.g., identity, log, logit).

Using the notation of a bold capital letter to represent a matrix, a bold lowercase greek letter to represent a parameter vector, the index $\bm{s}$ to represent a point in space (x and y and coordinates), and the index $t$ to represent a point in time,
the general form of a GLMM is
$g(\bm{\mu}) = \bm{X \beta} + \bm{Z \psi}$,
with the inverse link function $g(x)$ relating the mean $\bm{\mu}$ to fixed-effect predictors in the design matrix $\bm{X}$ with estimated coefficients $\bm{\beta}$ and random effects with the design matrix $\bm{Z}$ with estimated coefficients $\bm{\psi}$.
<!-- The matrix $\bm{Z}$ allows random effects to be structured (e.g., intercepts varying by group) or unstructured, with $\bm{u}$ representing random variation. -->
For clarity, we partition the random effects $\bm{Z \psi}$ into a number of more interpretable components, each of which can optionally be included in the model. 
Although in practice it is rare to include all of these components, the full sdmTMB model is
$$
g(\mu_{\bm{s},t}) =
\bm{X}_{1,\bm{s},t} \bm{\beta} +
\bm{\alpha}_g +
\bm{X}_{2,\bm{s},t} \gamma_t +
\bm{X}_{3,t} \zeta_{\bm{s}} +
\omega_{\bm{s}} +
\epsilon_{\bm{s},t},
$$
where

@@

Models constructed with sdmTMB may be viewed as spatially explicit models---they estimate interpretable parameters of a spatial covariance function: parameters defining the magnitude of spatial variation and the rate of correlation decay with distance.
In contrast, semi- and non-parametric approaches do not estimate spatial covariance functions (e.g., randomForest [@liaw2002], MaxEnt [@phillips2006], most smooths in mgcv [@wood2017a]).
The random fields in sdmTMB are structured as MVN constrained by a Matérn covariance function [@matern1960].
The Matérn can accommodate a range of shapes and can be both isotropic (covariance decays the same in all directions) or anisotropic (covariance in the latitudinal direction may differ from the longitudinal direction).
The Matérn standard deviations are estimated separately for the various components and the range can be shared or estimated separately.

@@

By default, we assume that if spatiotemporal fields are included, they are independent and identically distributed (IID); however, additional options allow them to be modelled as a random walk or first-order autoregressive, AR(1), process (Fig.\ \@ref(fig:diagram)).
Turning off both spatial and spatiotemporal effects allows comparison with a standard non-spatial GLM or GLMM.
We include additional flexibility in specifying the linear fixed effect matrix: covariates can be modelled as penalized smooth functions (generalized additive models; GAMs) using the same `s()` syntax as in mgcv [@wood2017a], or to be modelled with threshold shapes [e.g., hockey stick models, `breakpt()`, @barrowman2000] (Fig.\ \@ref(fig:diagram)).

The sdmTMB model is fit by maximum marginal likelihood.
Internally, a TMB model template calculates the marginal log likelihood and its gradient, and the negative log likelihood is minimized via the optimization routine `stats::nlminb()` in R (REF).
Random effects are estimated at values that maximize the log likelihood conditional on the estimated fixed effects and are integrated over via the Laplace approximation (REF).

@@

@@

Validation of mixed-effects models is non-trivial (that REF).

We provide three tools to assist with model validation:

1. The `residuals()` method returns randomized quantile residuals (e.g., as in DHARMa REF).
These residuals are calculated as $\hat{y} - y$ where $\hat{y}$ is the sum of the fixed effect estimates and the random effect values that maximumize the likelihood conditional on the estimated fixed effects. These residuals have known statistical issues (REF). One-step-ahead residuals are also possible or residuals from a single MCMC draw, but both approaches are much slower.

@@

3. An sdmTMB model can be passed to the tmbstan package [@monnahan2018] to sample from the joint posterior with Stan (REF) to evaluate the accuracy of the Laplace approximation. 
We note this will be considerably slower to fit than with maximum likelihood and in some cases may require tight priors or fixed parameters to ensure adequate sampling (REF).

Following glmmTMB, we allow AIC to be calculated for each sdmTMB model [@akaike1974]. 
Because of well-documented biases of AIC with mixed-effects models [@liang2008] we also provide functionality to perform estimation via k-fold cross-validation. 
Our function `sdmTMB_cv()` allows for any number of folds, with assignments being user-specified or chosen randomly. 
The predictive log-density is calculated for each fold and summarized across folds for a total measure of the predictive accuracy of a model (ELPD; REF).

@@

sdmTMB models can include penalized likelihoods by assigning priors to model parameters (`?sdmTMBpriors`).
These priors may be useful in cases where estimation is difficult because of identifiability issues, relatively flat likelihood surfaces, to impart prior information, or to achieve regularization.
Following other recent SPDE implementations [@simpson2017; @osgood-zimmerman2021], penalized complexity priors (`pc_matern`) may be particularly useful to constrain the spatial range and variance parameters.

@@

The results from a model fit with sdmTMB can be used to generate various quantities of interest.
For example, sdmTMB provides functionality for generating population-level summaries for each time slice including the center of gravity (`get_cog()`) and total population index (`get_index()`).
Estimated parameters from the Matérn covariance function can be used to describe the marginal spatial variance and Matérn range (distance at which spatial correlation decays to $\sim$ 0.13).

@@

An sdmTMB model requires a data frame that contains a response column, columns for any predictors, and columns for spatial coordinates.
It usually makes sense to convert the spatial coordinates to UTMs such that distance remains constant throughout the study region (e.g., using `sf::st_transform()`).
Here, we illustrate a spatial model fit to Pacific cod (*Gadus macrocephalus*) trawl survey data from Queen Charlotte Sound, BC, Canada.
Our model contains a main effect of depth as a penalized smoother, a spatial random field, and Tweedie observation error.
Our data frame `pcod` (built into the package) has a column `year` for the year of the survey, `density` for density of Pacific Cod in a given survey tow, `depth` for depth in meters of that tow, and spatial coordinates `X` and `Y`, which are UTM coordinates in kilometers:

@@

```{r, echo=TRUE, eval=FALSE}
head(pcod)
```

@@

The output indicates our model was fit by maximum (marginal) likelihood (`'ML'`).
We also see the formula, mesh, fitted data, and family.
Next we see any estimated fixed effects, the Tweedie dispersion and power parameters, the Matérn range distance, the marginal spatial field standard deviation, and the negative log likelihood at convergence.

We can make predictions with the `predict()` method; we will predict on new data with the `newdata` argument.
The data frame `qcs_grid` (included in the package data) is a grid that extends over the full survey domain and contains all predictors used in the model set at values for which we we want to predict.
The output of `predict()` is a data frame containing overall estimates, estimates from the non-random-field components (intercept and depth), and estimates from the random field components.
We show a basic plot of the estimated spatial random field, predictions across a depth gradient, and predictions in space in Fig. \@ref(fig:pcod-fig)b--d.

@@

There are many R packages available for fitting spatially explicit models to large datasets and we cannot review them all here [see @heaton2017]; however, the two most closely related are VAST and inlabru. 
VAST is also built on TMB and, for univariate data, should be able to fit equivalent models with some minor differences in functionality.
The difference lies mainly in the user interfaces and considerable additional functionality for multivariate response data and delta/hurdle models in VAST.
Interface-wise, VAST is a less modular interface (e.g., automatically constructing meshes before fitting, making predictions at the same time as fitting), is written from a fisheries index standardization perspective (e.g., splitting 'catchability' and 'habitat' covariates), and lacks some of the features that we think make sdmTMB more broadly usable by applied ecologists. 
For example, `sdmTMB()` has a `family` argument as in `glm()`, a `predict()` method, and human-readable options for spatial/spatiotemporal fields (e.g., `spatial = 'on'`, `spatiotemporal = 'AR1`).
However, for users who require multivariate spatiotemporal models or the Poisson-link delta model (REF), there is no substitute for VAST.

inlabru is similar to sdmTMB in design, but differs fundamentally in that fitting is achieved via INLA and approximate Bayesian computation.
For some models, TMB/sdmTMB can be considerably faster than INLA/inlabru [@osgood-zimmerman2021], but in our experience, those differences will depend on many aspects of the data, SPDE mesh, model, and fitting algorithm settings (Supp fig?).
One major capability included in sdmTMB but not INLA/inlabru is anisotropy.
We think the sdmTMB interface will be more familiar than inlabru to users of `glm()`, lme4, mgcv, or glmmTMB, and sdmTMB works with regular R data frames whereas inlabru works with `sp` class data objects.
However, inlabru fits additional model classes including log Gaussian Cox processes and may be particularly usable to those already familiar with INLA.

mgcv, but, yikes, do we have space:

- mgcv is capable of fitting with the SPDE (REF)
- typical smoothers don't offer MVN interpretability
- Tweedie is a big factor... must fix power 'p' to fit a GAMM.
- Barrier approaches are different
- doesn't have functionality built in for indexes/COG and uncertainty on such derived parameters
- for this class of models, we think sdmTMB has advantages, but mgcv is an immensely powerful modelling package that extends well beyond spatial/spatiotemporal data.

Final paragraph touching on other stuff possible in one paragraph: priors, Bayes, barrier meshes, randomized quantile residuals, simulation from joint precision, ...?

@@

# Application Examples

@@

# Tables

@@

>>>>

sdmTMB fits spatially explicit geostatistical GLMMs to point data.
Point data simply means the data have explicit spatial coordinates (vs. areal data that might represent polygons or grids).
The process component of an sdmTMB model can be formed by any combination of main ('fixed') effects, spatial random fields, spatiotemporal random fields, IID random intercepts, time-varying effects, and spatially varying effects (Fig. \@ref(fig:diagram)).
This process component is combined with an observation error family (e.g., Gaussian, Gamma, binomial, Tweedie) and link (e.g., identity, log, logit).

Using the notation of a bold capital letter to represent a data vector, a bold lowercase greek letter to represent a parameter vector, the index $\bm{s}$ to represent a point in space, and the index $t$ to represent a point in time,
the general form of a GLMM is
$g(\bm{\mu}) = \bm{X \beta} + \bm{Z \psi}$,
with the inverse link function $g(x)$ relating the mean $\bm{\mu}$ to fixed-effect predictors in the design matrix $\bm{X}$ with estimated coefficients $\bm{\beta}$ and random effects with the design matrix $\bm{Z}$ with estimated coefficients $\bm{\psi}$.
<!-- The matrix $\bm{Z}$ allows random effects to be structured (e.g., intercepts varying by group) or unstructured, with $\bm{u}$ representing random variation. -->
For clarity, we partition the random effects $\bm{Z \psi}$ into a number of more interpretable components, each of which can optionally be included in the model. 
Although in practice it is rare to include all of these components, the full sdmTMB model is
$$
g(\mu_{\bm{s},t}) =
\bm{X}_{1,\bm{s},t} \bm{\beta} +
\bm{\alpha}_g +
\bm{X}_{2,\bm{s},t} \gamma_t +
\bm{X}_{3,t} \zeta_{\bm{s}} +
\omega_{\bm{s}} +
\epsilon_{\bm{s},t},
$$
where

@@

sdmTMB models are spatially explicit models---they estimate interpretable parameters of a spatial covariance function: parameters defining the magnitude of spatial variation and the rate of correlation decay with distance.
In contrast, semi- and non-parametric approaches do not estimate spatial covariance functions (e.g., randomForest [@liaw2002], MaxEnt [@phillips2006], most smooths in mgcv [@wood2017a]).
The random fields in sdmTMB are structured as MVN constrained by a Matérn covariance function [@matern1960].
The Matérn can accommodate a range of shapes and can be both isotropic (covariance decays the same in all directions) or anisotropic (covariance in the latitudinal direction may differ from the longitudinal direction) (REF).
The Matérn standard deviations are estimated separately for the various fields and the range---the distance at which spatial correlation decays to $\sim$ 0.13 (REF)---can be shared or estimated separately.

@@

By default, if spatiotemporal fields are included, they are assumed independent and identically distributed (IID); however, additional options allow them to be modelled as a random walk or first-order autoregressive, AR(1), process (Fig.\ \@ref(fig:diagram)).
Turning off both spatial and spatiotemporal effects allows comparison with a standard non-spatial GLM or GLMM.
We include additional flexibility in specifying the linear fixed effect matrix: covariates can be modelled as penalized smooth functions (generalized additive models; GAMs) using the same `s()` syntax as in mgcv [@wood2017a], or to be modelled with threshold shapes [e.g., hockey stick models, `breakpt()`, @barrowman2000] or logistic functions, `logistic()`.

The sdmTMB model is fit by maximum marginal likelihood.
Internally, a TMB [@kristensen2016] model template calculates the marginal log likelihood and its gradient, and the negative log likelihood is minimized via the non-linear optimization routine `stats::nlminb()` in R [@gay1990; @r2021].
Random effects are estimated at values that maximize the log likelihood conditional on the estimated fixed effects and are integrated over via the Laplace approximation [@kristensen2016].

@@

Following glmmTMB, we allow AIC to be calculated for each sdmTMB model [`aic()`, @akaike1974]. 
Because of well-documented biases of AIC with mixed-effects models [@liang2008] we provide functionality to perform k-fold cross-validation.
The function `sdmTMB_cv()` allows for any number of folds, with assignments being user-specified or chosen randomly. 
The predictive log-density is calculated for each fold and summarized across folds for a total measure of the predictive accuracy of a model.

@@

Validation of mixed-effects models is non-trivial (that REF). We provide three approaches to assist with model validation:

1. The `residuals()` method returns randomized quantile residuals (e.g., as in DHARMa REF).
These residuals are calculated as $\hat{y} - y$ where $\hat{y}$ is the sum of the fixed effect estimates and the random effect values that maximumize the likelihood conditional on the estimated fixed effects. 
These residuals have known statistical issues (REF) but are quick to calculate. One-step-ahead residuals (REF) or residuals from a single MCMC draw (REF) are also possible, but both approaches are much slower.

@@

3. An sdmTMB model can be passed to the tmbstan package [@monnahan2018] to sample from the joint posterior with Stan (REF) to evaluate the accuracy of the Laplace approximation (see `?extract_mcmc`). 
We note this will be considerably slower to fit than with maximum likelihood and in some cases may require tight priors or fixed parameters to ensure adequate sampling (REF).

@@

sdmTMB models can include penalized likelihoods by assigning priors to model parameters (`?sdmTMBpriors`).
These priors may be useful in cases where estimation is difficult because of identifiability issues or relatively flat likelihood surfaces, or to impart prior information or achieve regularization.
Following other recent SPDE implementations [@simpson2017; @osgood-zimmerman2021], penalized complexity priors (`pc_matern()`) may be particularly useful to constrain the spatial range and variance parameters.

@@

The results from an sdmTMB model can be used to generate various quantities of interest.
For example, sdmTMB provides functionality for generating population-level summaries for each time slice including the center of gravity (`get_cog()`) and total population index (`get_index()`).
Estimated parameters from the Matérn covariance function can be used to describe the marginal spatial variance and Matérn range.

@@

An sdmTMB model requires a data frame that contains a response column, columns for any predictors, and columns for spatial coordinates.
It usually makes sense to convert the spatial coordinates to UTMs such that distance remains constant throughout the study region (e.g., using `sf::st_transform()`).
Here, we illustrate a spatial model fit to Pacific cod (*Gadus macrocephalus*) trawl survey data from Queen Charlotte Sound, BC, Canada.
Our model contains a main effect of depth as a penalized smoother, a spatial random field, and Tweedie observation error.
Our data frame `pcod` (built into the package) has a column `year` for the year of the survey, `density` for density of Pacific cod in a given survey tow, `depth` for depth in meters of that tow, and spatial coordinates `X` and `Y`, which are UTM coordinates in kilometers:

@@

```{r, echo=TRUE, eval=FALSE}
library(sdmTMB)
head(pcod)
```

@@

The output indicates our model was fit by maximum (marginal) likelihood (`'ML'`).
We also see the formula, mesh, fitted data, and family.
Next we see any estimated fixed effects, the Tweedie dispersion and power parameters, the Matérn range distance, the marginal spatial field standard deviation, and the negative log likelihood at convergence.
Internally, `s(depth)` creates random effect spline weights and so is not shown here.

We can make predictions with the `predict()` method; we will predict on new data with the `newdata` argument.
The data frame `qcs_grid` (included in the package data) is a grid that extends over the full survey domain and contains all predictors used in the model set at values for which we want to predict.
The output of `predict()` is a data frame containing overall estimates, estimates from the non-random-field components (intercept and depth), and estimates from the random field components.
We show a basic plot of the estimated spatial random field, predictions across a depth gradient, and predictions in space in Fig. \@ref(fig:pcod-fig)b--d.

@@

There are many R packages available for fitting spatially explicit models to large datasets and we cannot review them all here [see @heaton2017]; however, the two most closely related are VAST and inlabru. 
VAST is also built on TMB and, for univariate data, should be able to fit equivalent models with some minor differences in functionality.
The difference lies mainly in the user interfaces and considerable additional functionality for multivariate response data (REF) and delta/hurdle models (REF) in VAST.
Interface-wise, VAST is a less modular interface (e.g., automatically constructing meshes before fitting, making predictions at the same time as fitting), is written from a fisheries index standardization perspective (e.g., splitting 'catchability' and 'habitat' covariates), and lacks some of the features that we think make sdmTMB more broadly usable by applied ecologists. 
For example, `sdmTMB()` has a `family` argument as in `glm()`, a `predict()` method, and human-readable options for spatial/spatiotemporal fields (e.g., `spatial = 'on'`, `spatiotemporal = 'AR1'`).
sdmTMB achieves 3 of the 4 design principles outlined in Thorson (REF) for index standardization: it allows area weighting, can distinguish catchability and habitat covariates (by deciding what is fixed when predicting), and conditions on missing covariates (random fields).
Unlike VAST, however, sdmTMB does not currently extend to multivariate applications.

<!--         -   "VAST represents four main design principles. Although there are many other implementation details that distinguish VAST from previous index-standardization methods, I believe that these four design principles represent the features that any future replacement should include." -->
<!--         -   Area weighting (check, helper functions to sum up weighted grid cells) -->
<!--         -   Distinct catchability and habitat covariates (check, decide what goes into `predict()`) -->
<!--         -   Condition on missing covariates (check, random fields) -->
<!--         -   Bridge between univariate and multivariate applications (no! should we? -->
<!-- However, VAST can fit multivariate spatiotemporal models (REF) and Poisson-link delta model (REF), whereas sdmTMB cannot. -->

inlabru is similar to sdmTMB in design, but differs fundamentally in that fitting is achieved via INLA and approximate Bayesian computation.
For some models, TMB/sdmTMB can be considerably faster than INLA/inlabru [@osgood-zimmerman2021], but in our experience, those differences depend on many aspects of the data, SPDE mesh, model, and fitting algorithm settings (TODO Supp fig?).
One major capability included in sdmTMB but not INLA/inlabru is anisotropy.
Furthermore, we think the sdmTMB interface will be more familiar than inlabru to users of `glm()`, lme4, mgcv, or glmmTMB.
On the other hand, inlabru fits additional model classes including log Gaussian Cox processes and may be particularly usable to those already familiar with INLA.

mgcv can fit many similar models to sdmTMB either with various flavours of splines or even with the SPDE approximation to random fields (REF); however, there are some notable advantages of sdmTMB for the specific class of models discussed here.
First, the smoothers typically used with mgcv (thin plate splines, ...), while capable of making equivalent predictions (REF) do not offer the ecological interpretability of a spatially explicit random field model (spatial range and variance).
Second, the implementation of the Tweedie distribution in mgcv requires the power parameter to be fixed in a mixed effects model.
Third, mgcv does not have built-in functionality to calculate derived values with uncertainty such as population indexes or center of gravity.
Despite these sdmTMB advantages for the models discussed here, mgcv is an immensely powerful modelling package that extends well beyond spatial random field models.

Spatially and spatiotemporally explicit data are increasingly collected in ecology and have the power to reveal new ecological processes (REF) and improve ecological management (REF).
These data present statistical challenges to modelling them effectively and efficiently; appropriate models such as GLMMs with random fields are often computationally intensive and challenging to implement, interpret, and evaluate.
We hope the development of user-friendly interfaces such as sdmTMB opens this useful class of models to a wider audience of users.

@@

@@

<!--
# Tables

@@

-->
